Laurentia's continued motion into the Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation

### 2022-07-17 Manuscript/Manuscript.tex

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca.990 Ma Jacobsville Formation 
\item Reversals put younger bounds on the normal Keweenawan superchron 
\item The new pole indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of the speed of plate motion is the onset of the Grenvillian Orogeny which lasted from ca. 1090-980 Ma. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum deposition age has been constrained to be ca. 990 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, USA. High-resolution thermal demagnetization experiments resolve primary magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, which provides constraints for the duration of the Keweenawan superchron. Inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province may record a coherent plate motion after major continental collision during the assemblage of supercontinent Rodinia rather than recording a significant crustal shortening.
\end{abstract}

@@

\begin{keypoints}
\item Neoproterozoic
\item Laurentia  
\item Paleogeography
\item Inclination shallowing
\item Grenville
\item Keweenawan superchron
\item Red bed
\end{keypoints}

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provides crucial constraints on global paleogeography in the Mesoproterozoic and has become the central record for reconstructing the development of ancient supercontinent Rodinia \cite{Swanson-Hysell2019a, Swanson-Hysell2021c}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a quiescence period that could have lasted as much as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. As a result, uncertainties exist in reconstructing Laurentia and its conjugate continents' configuration in the late Mesoproterozoic to early Neoproterozoic.

@@

\subsection*{Fluvial intraclast conglomerate test}

@@

@@

\section*{Discussion}
\subsection*{Inclination shallowing correction and Jacobsville paleomagnetic pole}

Although igneous and sedimentary rocks associated with the North America Midcontinent Rift provided high-resolution record of the ca. 1110 to 1070 Ma Keweenawan Track which is the apparent polar wander path (APWP) for Laurentia in the Mesoproterozoic, high-quality paleogeographic constraints is lacking thereafter since tectonism in the interior of Laurentia entered a protracted period of quiescence until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. The new paleomagnetic data developed from the Jacobsville Formation presents an opportunity to extend the Keweenawan Track toward the early Neoproterozoic as its maximum deposition age has been recently tightly constrained to be ca. 992 Ma \cite{Hodgin2022a}.

However, inclination bias associated with sedimentary rocks need to be corrected when synthesizing a paleomagnetic pole. Rotation of ferromagnetic grains during sediment deposition and compaction during early diagenesis can result in the acquisition of a detrital remanent magnetization that is biased shallow relative to the local geomagnetic field in which it was acquired \cite{King1955a, Tauxe2005a, Kodama2012a}. This effect can be summarized as: 
\begin{equation*}
\textup{tan}(I_o) = f\textup{tan}(I_f)
\end{equation*}
where $f$ represents the amount of inclination shallowing, $I_o$ represents the observed inclination of the specimen magnetization and $I_f$ represents the inclination of the field in which the magnetization was acquired. If uncorrected, shallower inclinations obtained from sedimentary rocks can result in erroneously low estimates of paleolatitudes, biasing the interpreted past positions of continents and hindering plate reconstructions \cite{Domeier2012a}. Methods for correcting sedimentary inclination bias have been developed. They include laboratory-based measurements of magnetic mineral fabrics \cite<e.g.>{Kodama1992a, Bilardello2015a}, as well as statistical methods based on the evaluation of the deviation of virtual geomagnetic pole (VGP) distribution patterns recorded by sedimentary rocks with an assumed paleosecular variation model (the ``elongation/inclination" [E/I] method of \citeA{Tauxe2004b}). When a large number of sedimentary paleomagnetic samples (typically n$>$100) capture an extensive period of time such that the paleosecular variation is sufficiently averaged, the E/I method can give accurate estimates and is more efficient than lab-based methods. In the ca. 1.1 Ga Midcontinent Rift, \citeA{Pierce2022a} showed that the E/I method is effective in correcting the inclination bias recorded by the Cut Face Creek Sandstone---a lava flow-bracketed fluvial red bed sedimentary formation dominated by overbank deposition.

Inclination shallowing is also present in the Jacobsville Formation. Scatter of the specimen high-temperature component directions in geographic coordinates show dispersion of declinations parallel to the bedding plane, indicating the directions are likely shallowed along the bedding-vertical plane (Fig. \ref{fig:in_situ_pmag}; \citeA{Tauxe2004b}). Since the deposition of the Jacobsville Formation span over geomagnetic field reversals, it is of interest to investigate the amount of inclination shallowing associated with the normal (n=130) and the reversed (n=148) polarity respectively. Applying the E/I method on the directions of normal polarity yields an estimated $f$ value of 0.67 with 95\% bootstrap uncertainty bounds from 0.91 to 0.55 (Fig. \ref{fig:in_situ_pmag}B). On the other hand, directions of reversed polarity yields an estimated $f$ factor of 0.52 with more uncertainties ranging from 0.97 to 0.29 (Fig. \ref{fig:in_situ_pmag}B). In addition, the dual-polarity directions do not pass a reversal test \cite{McFadden1990a, Tauxe2010a}, regardless of whether inclination correction is applied (Fig. \ref{fig:in_situ_pmag}).

One interpretation of the polarity discrepancy could be that the deposition of the Jacobsville Formation was protracted such that the sediments at Hungarian Falls which dominantly record a reversed polarity was deposited prior to those at Baby Snake Creek which dominate the normal polarity records (Fig. \ref{fig:in_situ_pmag}). This is consistent with the magnetostratigraphic interpretation that Baby Snake Creek section is younger than the Hungarian Falls section (Fig. \ref{fig:strat_column}). Field observations of \citeA{Tyrrell2019a} and \citeA{Mueller2021a} near Bete Grise Bay, which is only $\sim$10 km east to Baby Snake Creek (Fig. \ref{fig:Geologic_map}), show that significant thrusting and deformation along the Keweenaw Fault and subsequent weathering of the Midcontinent Rift basalt occurred prior to the deposition of the Jacobsville Formation. In addition, U-Pb calcite geochronology and thermometry confirm that the final motion of $\sim$2 km of vertical displacement along the Keweenaw Fault occurred during the ca. 980 Ma Rigolet phase of the Grenvillian Orogeny and the majority of the Keweenaw Fault motion likely occurred much earlier during the ca. 1050 Ma Ottawan phase orogeny \cite{Hodgin2022b}. Therefore, we interpret that it is most likely that the Jacobsville Formation were deposited just prior to and during the Rigolet phase of the Grenvillian Orogeny with progressive sediment transportation to the northwest resulting in its onlapping on top of the volcanics in the northern tip of the Keweenaw Peninsula. As a result, the sediments at Baby Snake Creek records a slightly more southerly paleolatitude than those near Hungarian Falls as Laurentia's plate motion continues toward the south (Fig. \ref{fig:in_situ_pmag}, \ref{fig:pole_plot}).

An alternative interpretation to the polarity discrepancy could be that there are insufficient samples such that either or both polarities are not average for secular variation.

Overall, both inclination-corrected polarities recorded by the Jacobsville Formation show mean pole positions in the southern hemisphere with respect to today's continental configuration (normal polarity longitude=, latitude=, A95=; reversed polarity longitude= , latitude= , A95=). The two pole positions are close to each other, but are distinct from the youngest poles developed from rocks of the Midcontinent Rift (Fig. \ref{fig:pole_plot}). In addition, albeit that the Jacobsville sedimentation span over geomagnetic field reversals, the large arc distance between the poles of the Jacobsville Formation and that of the Freda Formation precludes the hypothesis that the Jacobsville Formation at the five studied localities (Fig. \ref{fig:Geologic_map}) is coeval with the Oronto Group and formed through a prolonged period of time.

To succinctly summarize a paleomagnetic pole for the ca. 992 Ma Jacobsville Formation, we combine the directions of both polarities and reassess the amount of inclination shallowing of the combined directional dataset (Fig. \ref{fig:in_situ_pmag}). The resultant estimate for the $f$ value is 0.61, and the inclination-corrected mean direction is dec= , inc= , $\alpha_{95}$= . The bootstrap 95\% uncertainty range for the $f$ value range from 0.80 to 0.51, which overlaps with the uncertainty ranges of directions associated with each polarity (Fig. \ref{fig:in_situ_pmag}). The Fisher mean paleomagnetic pole associated with the combined directions is shown in Fig. \ref{fig:pole_plot} (pole longitude= 186.3\textdegree E, pole latitude= 14.3\textdegree S, A95= 2.5\textdegree).

\citeA{Pierce2022a} discussed that as sedimentary grains usually flatten perpendicular to the bedding plane but circularly symmetric about the bedding plane, the resultant shallow inclinations they record would typically result in asymmetric uncertainties in the recorded paleomagnetic pole position with more uncertainties in paleolatitues than paleolongitudes. The asymmetric uncertainties associated with sedimentary paleomagnetic poles violate the assumption of a Fisher distribution \cite{Fisher1953a} where uncertainties associated with a mean pole position is circularly distributed. Instead, the spherical asymmetric bivariate Kent distribution \cite{Kent1983a} is more appropriate for incorporating uncertainties associated with inclination shallowing in reported sedimentary paleomagnetic poles. We follow the approach of \citeA{Pierce2022a} who extended the E/I method of \citeA{Tauxe2004b}. First we take all estimated $f$ values from the bootstrap results of the E/I algorithm and apply each value to correcting all Jacobsville detrital remanence directions. Then we calculate the Fisher statistics for each of the inclination-corrected mean poles. Finally we obtain 100 Monte Carlo resamples of each Fisher mean pole. As a result, the distribution of the resampled mean pole positions have an elongated distribution which represents more uncertainty associated with paleolatitudes of Laurentia than paleolongitudes (Fig. \ref{fig:pole_plot}). The distribution of paleolatitudes of the resampled mean poles can be well-approximated by a normal distribution (Fig. \ref{fig:pole_plot}). We use the Kent distribution to parameterize the 95\% confidence ellipse of the distribution of the resampled mean poles (mean longitude=186.4\textdegree E, mean latitude=14.2\textdegree S, major axis longitude=262.5, major axis latitude=43.5, major axis magnitude=3.3\textdegree, minor axis longitude=110.0, minor axis latitude=43.0, minor axis magnitude=3.0\textdegree; Fig. \ref{fig:pole_plot}). As is shown in Fig. \ref{fig:pole_plot}, although the Kent ellipse uncertainty range of the Jacobsville Formation's paleomagnetic pole position has significant overlap with the circular uncertainty range represented by the Fisher mean, the Kent ellipse informs more uncertainty along the great circle path between the mean pole position and the locality of the Jacobsville Formation as it incorporates the uncertainties in the amount of inclination shallowing within the sediments.

% \subsection*{Age of the Grenville Loop}

\subsection*{Paleogeography of Laurentia during the late Mesoproterozoic to early Neoproterozoic}

\citeA{McCabe1983a} developed paleomagnetic data from the Chequamegan Sandstone of the Bayfield Group and found reversed-polarity directions similar to that of the Jacobsville Formation poles developed by \citeA{Roy1978a}.

Although igneous paleomagnetic records from the the interior of Laurentia has been lacking since the end of the Midcontinent Rift magmatism, data from metamorphosed rocks associated with the ca. 1080 to 980 Ma Grenville continental collisional orogenesis has been developed to infer for Laurentia's plate configuration in the late Mesoproterozoic to early Neoproterozoic \cite<e.g.>{Weil1998a, Halls2015b}. However, in contrast to igneous rocks of the Midcontinent Rift which have relatively simple thermal history and straightforward paleomagnetic records \cite,e.g.>{Swanson-Hysell2019a}, rocks of the Grenville Province which experienced up to granulite facies metamorphism have much more complicated thermal history and the timing of their magnetic remanence acquisition is more difficult to constrain. As is shown in Fig. \ref{fig:pole_plot}, paleomagnetic poles developed from rocks of the Grenvillian Orogeny plots near Australia in today's continental reference frame, forming an arc distance of $\sim$30 degrees from the youngest paleomagnetic poles of the ca. 1070 Ma Freda Formation \cite{Henry1977a}. Such great discrepancies between pole positions associated with uncertainties in the age of the magnetization of the Grenville rocks led to the ``Grenville Problem" which is still highly debated today - did the Grenville rocks acquire magnetic remanence prior to the continental collision and thus the pole discrepancies are a result of significant crustal shortening (4000 $\pm$ 1000 km; \citeA<e.g.>{Halls2015b}), or did they acquire remanence after the continental accretion and thus the southerly poles merely represent Laurentia's continued motion through more recent times \cite<e.g.>{Weil1998a}?

As the ca. 990 Ma Jacobsville Formation was deposited in Laurentia Interior (Fig. \ref{fig:Geologic_map}), its paleomagnetic pole position can help test the contrasting hypothesis regarding the ``Grenville Problem".

paleomagnetic mean pole of the Jacobsville Sandstone and the age

inclination shallowing correction of the Jacobsville Sandstone

magnetic reversal during the Jacobsville and the Maya Superchron

Implication on the crustal shortening of the Rigolet phase of the Grenville Orogeny

Warnock on the age of magnetization on the Haliburton pole: based on the very coarse thermal demagnetization steps Warnock2000a assigned that all of the primary magnetite components unblocked above 550 \textdegree C, with an uncertainty of 555 \textdegree C $\pm$ 5\textdegree C. Using a thermal history model that combines U-Pb titanite ages, hornblende and biotite Ar/Ar ages and just by merely compare the thermal history model with the assigned narrow magnetite unblocking temperature range, the authors assigned the age of the magnetization to be 1015 $\pm$ 15 Ma.

It is important to note that the result Ar ages of the hornblende and biotite are 983 $\pm$ 8 Ma, 996 $\pm$ 8 Ma, and 998 $\pm$ 8 for hornblende with unblocking temperature of 530 $\pm$ 40\textdegree C, consistent across metamorphic terranes (Bancroft). And the biotite Ar ages are 910 $\pm$ 7 Ma, 891 $\pm$ 7 Ma, corresponding to unblocking temperature of 280 $\pm$ 40\textdegree C.

More, the thermal demagnetization plots shown in Warnock2000a suggest that the magnetite unblocking lower temperature boundary can be around 500\textdegree C, lower than 555 $\pm$ 5\textdegree C. In addition, given the case that the rocks underwent uniform rate of cooling due to post-formation uplifting, a slow cooling effect from 555\textdegree C to 300\textdegree C from 990 Ma to 900 Ma allows for a scenario where the magnetite remament magnetization that appear to have a 555\textdegree C unblocking temperature in lab was acquired through a period of --- myr cooling though 500\textdegree C during uplifting. In this case, the magnetite remanence acquisition can be interpreted to be closer to the biotite Ar age than previously thought. Therefore, the Haliburton intrusion pole could be interpreted to be a post Rigolet orogenic activity phase magnetization acquisition, and the apparent discrepancy between the Haliburton pole and the Jacobsville pole can be explained by Laurentia's whole plate motion after the Grenvillian orogeny from 990 Ma to 900 Ma. One corollary of such scenario is that Laurentia underwent plate motion at a rate of ---\textdegree/myr, which is similar to that of today's tectonic plate motion.

Crucial link between the Keweenawan Track and the Grenville Loop

>>>>

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{amsmath}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca. 992 Ma Jacobsville Formation 
\item Reversals put younger bounds on the normal Keweenawan superchron 
\item The new pole indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of Laurentia's plate speed is the onset of the ca. 1090-980 Ma Grenvillian Orogeny along the eastern margin. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum deposition age has been constrained to be ca. 992 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along 5 stratigraphic sections in northern Michigan, USA. High-resolution thermal demagnetization experiments resolve detrital magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, providing constraints for the duration of the Keweenawan superchron. Inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the development of supercontinent Rodinia.
% rather than recording a significant crustal shortening.
\end{abstract}

@@

\begin{keypoints}
\item Neoproterozoic
\item Laurentia  
\item Paleogeography
\item apparent polar wander path
\item Inclination shallowing
\item Grenville
\item Keweenawan superchron
\item Red bed
\end{keypoints}

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provides crucial constraints on global paleogeography in the Mesoproterozoic and has become a central record for reconstructing the development of ancient supercontinent Rodinia \cite{Swanson-Hysell2019a, Swanson-Hysell2021c}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a quiescence period that could have lasted as much as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. As a result, uncertainties exist in reconstructing Laurentia and its conjugate continents' configuration in the late Mesoproterozoic to early Neoproterozoic.

@@

\subsection*{Paleomagnetic field test}

\subsubsection*{Fluvial intraclast conglomerate test}

@@

\subsubsection*{Fold test}

@@

\section*{Discussion}
\subsection*{Inclination shallowing correction}

Although igneous and sedimentary rocks associated with the North American Midcontinent Rift provided high-resolution record of the ca. 1110 to 1070 Ma Keweenawan Track which is the apparent polar wander path (APWP) for Laurentia in the late Mesoproterozoic, high-quality paleogeographic constraints is lacking thereafter since magmatism in the interior of Laurentia entered a protracted quiescence period until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a}. The new paleomagnetic data developed from the Jacobsville Formation presents an opportunity to extend the Keweenawan Track toward the early Neoproterozoic as its maximum deposition age has been tightly constrained to be ca. 992 Ma through high-precision U-Pb detrital zircon geochronology \cite{Hodgin2022a}.

However, inclination bias associated with sedimentary paleomagnetic directions need to be corrected when synthesizing poles. Rotation of grains during sediment deposition and compaction during early diagenesis can result in the acquisition of a detrital remanent magnetization that is more shallowly inclined relative to the local geomagnetic field in which it was acquired \cite{King1955a, Tauxe2005a, Kodama2012a}. This effect can be summarized as: 
\begin{equation*}
\textup{tan}(I_o) = f\textup{tan}(I_f)
\end{equation*}
where $f$ represents the amount of inclination shallowing, $I_o$ represents the observed inclination, and $I_f$ represents the inclination of the field in which the magnetization was acquired. If uncorrected, shallower inclinations obtained from sedimentary rocks can result in erroneously low estimates of paleolatitudes, biasing paleogeographic reconstructions \cite{Domeier2012a}. Methods for correcting sedimentary inclination bias include laboratory-based measurements of magnetic mineral fabrics \cite<e.g.>{Kodama1992a, Bilardello2015a}, as well as statistical methods based on the evaluation of the deviation of observed virtual geomagnetic pole (VGP) distribution patterns from an assumed paleosecular variation model (the ``elongation/inclination" [E/I] method of \citeA{Tauxe2004b}). When a large number of sedimentary paleomagnetic samples (typically n$>$100) capture an extensive period of time such that the paleosecular variation is sufficiently averaged, the E/I method can give accurate estimates and is less labor-intensive than methods based on mineral fabric measurements. In the ca. 1.1 Ga Midcontinent Rift, \citeA{Pierce2022a} showed that the E/I method is effective in correcting the inclination bias recorded by a package of lava flow-bracketed fluvial red bed sediments---the Cut Face Creek Sandstone.

Inclination shallowing is also present in the fluvial red beds of the Jacobsville Formation. Scatter of the specimen detrital remanence directions in geographic coordinates show dispersion of remanence vectors parallel to the bedding plane, indicating the directions are likely shallowed along the bedding-perpendicular plane (Fig. \ref{fig:in_situ_pmag}; \citeA{Tauxe2004b}). Since the deposition of the Jacobsville Formation span over geomagnetic field reversals, it is of interest to investigate the amount of inclination shallowing associated with the normal (n=130) and the reversed (n=148) polarity respectively. Applying the E/I method on the normal polarity directions yields an estimated $f$ value of 0.67 with 95\% bootstrap uncertainty bounds from 0.91 to 0.55 (Fig. \ref{fig:in_situ_pmag}B). On the other hand, directions of reversed polarity yields an estimated $f$ factor of 0.52 with more uncertainties ranging from 0.97 to 0.29 (Fig. \ref{fig:in_situ_pmag}B). Although the estimated $f$ value for the reversed-polarity directions has large statistical uncertainty bounds, the uncertainties are likely associated with the overall shallow inclinations and distributions of directions in both upper and lower hemispheres (Fig. \ref{fig:in_situ_pmag}B). Therefore, inclination corrections do not result in significant changes in the mean reversed direction. In addition, directions of the two polarities do not pass a reversal test \cite{McFadden1990a, Tauxe2010a}, regardless of whether inclination correction is applied.

One interpretation of the polarity discrepancy could be that the deposition of the Jacobsville Formation was protracted such that the sediments at Hungarian Falls whose detrital remanence dominate reversed polarity records was deposited prior to those at Baby Snake Creek which dominate the normal polarity records (Fig. \ref{fig:in_situ_pmag}). This is consistent with the magnetostratigraphic interpretation that Baby Snake Creek section is younger than the Hungarian Falls section (Fig. \ref{fig:strat_column}). Field observations of \citeA{Tyrrell2019a} and \citeA{Mueller2021a} near Bete Grise Bay, which is only $\sim$10 km east to Baby Snake Creek (Fig. \ref{fig:Geologic_map}), show unconformable onlap of the Jacobsville Formation on top of Midcontinent Rift volcanics that have experienced significant thrusting, deformation. At some locations, paleosol development has been found on top of the Midcontinent Rift basalt along the segmented Keweenaw Fault branches prior to the deposition of the Jacobsville sediments. This is in contrast with the faulted contact between the Jacobsville Formation and the volcanics to the south where the Jacobsville Formation is often folded against the volcanics (Fig. \ref{fig:Geologic_map}). U-Pb calcite geochronology and thermometry confirm that the final $\sim$2 km vertical displacement along the Keweenaw Fault occurred during the ca. 980 Ma Rigolet phase of the Grenvillian Orogeny whereas the majority of the fault motion likely occurred during the ca. 1050 Ma Ottawan phase \cite{Hodgin2022b}. We interpret that it is most likely that deposition of the Jacobsville Formation initiated close to the Rigolet phase of the Grenvillian Orogeny with progressive sediment transportation to the northwest as the Grenville back-bulge basin propagated toward the interior of Laurentia \cite{Hodgin2022b}. As a result, we hypothesize that the sediments at Baby Snake Creek likely records a slightly younger pole position with a more southerly paleolatitude than those near Hungarian Falls, indicating Laurentia's plate motion toward the south (Fig. \ref{fig:EI_results}). However, we cannot rule out an alternative interpretation that the sediments at Baby Snake Creek are coeval or older than some of the sediments to the south, as the basal Jacobsville Formation at Baby Snake Creek is folded against the Midcontinent Rift volcanics. Further field mapping and geochronological constraints are needed to test this hypothesis.

% An alternative interpretation to the polarity discrepancy could be that there are insufficient samples such that either or both polarities are not average for secular variation. 
Overall, both inclination-corrected polarities recorded by the Jacobsville Formation show mean pole positions in the southern hemisphere with respect to today's continental configuration (normal polarity longitude=175.1\textdegree E, latitude=19.1\textdegree S, A95=3.8\textdegree; reversed polarity longitude=191.1\textdegree E, latitude=14.5\textdegree S, A95=3.9\textdegree; both are corrected for inclination shallowing; Fig. \ref{fig:EI_results}). The two pole positions are distinct from the youngest poles developed from rocks of the Midcontinent Rift (Fig. \ref{fig:EI_results}). Despite that the Jacobsville sedimentation span over geomagnetic field reversals, the large arc distance between either polarity of the Jacobsville Formation and the pole of the ca. 1070 Ma upper Freda Formation is consistent with the interpretation that the onset of deposition of the Jacobsville Formation at the five studied localities post-date the youngest member of the Oronto Group sediments by tens of millions of years.

\subsection*{The Jacobsville paleomagnetic pole}
To succinctly summarize a paleomagnetic pole for the ca. 992 Ma Jacobsville Formation, we combine the directions of both polarities and reassess the amount of inclination shallowing of the combined directional dataset (Fig. \ref{fig:in_situ_pmag}). The resultant estimate for the $f$ value is 0.61 with a bootstrap 95\% uncertainty range for the $f$ value between 0.80 to 0.51, which overlaps with the uncertainty ranges of directions associated with either polarity (Fig. \ref{fig:in_situ_pmag}). The $f$ value of 0.61 is strikingly close to the value of 0.59 which is a calculated mean value of a compilation of inclination shallowing factor for hematite-bearing rocks derived by \cite{Bilardello2010b}. The Fisher mean paleomagnetic pole associated with the combined directions is shown in Fig. \ref{fig:EI_results} (pole longitude= 186.3\textdegree E, pole latitude= 14.3\textdegree S, A95= 2.5\textdegree).

\citeA{Pierce2022a} discussed that as inclination shallowing processes usually do not bias the recorded declination, it would typically result in asymmetric uncertainties in the recorded paleomagnetic pole position with more uncertainties in paleolatitues than paleolongitudes. The asymmetric uncertainties in the mean pole position thus violate the assumption of the Fisher distribution \cite{Fisher1953a} where uncertainties are circularly distributed about a mean vector. Instead, the spherical asymmetric bivariate Kent distribution \cite{Kent1983a} is more appropriate for incorporating uncertainties associated with inclination shallowing in reported sedimentary paleomagnetic poles. We follow the approach of \citeA{Pierce2022a} who extended the E/I method of \citeA{Tauxe2004b}. First we take all estimated $f$ values from the bootstrap results of the E/I algorithm and apply each value to correct all Jacobsville detrital remanence directions (with the reversed directions flipped to antipodes). Then we calculate the Fisher statistics for each of the inclination-corrected mean pole. Finally, we perform a Monte Carlo approach to generate 100 resamples of each Fisher mean pole. As a result, the distribution of the resampled mean pole positions have an elongated distribution which represents more uncertainty associated with paleolatitudes of Laurentia than paleolongitudes (Fig. \ref{fig:EI_results}). The distribution of paleolatitudes of the resampled mean poles can be well-approximated by a normal distribution (Fig. \ref{fig:EI_results}). We use the Kent distribution to parameterize the 95\% confidence ellipse of the distribution of the resampled mean poles (mean longitude=186.4\textdegree E, mean latitude=14.2\textdegree S, major axis longitude=262.5, major axis latitude=43.5, major axis magnitude=3.3\textdegree, minor axis longitude=110.0, minor axis latitude=43.0, minor axis magnitude=3.0\textdegree; Fig. \ref{fig:EI_results}). As is shown in Fig. \ref{fig:EI_results}, although the Kent uncertainty ellipse has significant overlap with the circular uncertainty range calculated by the Fisher statistics due to the overall low paleolatitude of the Jacobsville basin, the Kent ellipse informs more uncertainty along the great circle path between the mean pole position and the locality of the basin as it incorporates the uncertainties in the amount of inclination shallowing within the sediments.

\citeA{Roy1978a} isolated the detrital magnetization component via alternating-field, thermal, and chemical ``cleaning" demagnetization on a suite of Jacobsville red beds near Keweenaw Peninsula (their area A), the town of Marquette (area B), and Sault Ste Marie (area C). Although least-squares principal component analyses was not a routine approach in fitting paleomagnetic directions at the time, that study had success in removing the present day local field overprint and was able to resolve a magnetic polarity transition at one locality. The mean paleomagnetic pole position calculated from the interpreted primary magnetic remanence from rocks of the Keweenaw Peninsula and Marquette (J1 A+B pole; pole longitude=183\textdegree E, pole latitude=9\textdegree S, dp=3\textdegree, dm=6\textdegree; no inclination correction; Fig. \ref{fig:EI_results}; \citeA{Roy1978a}) is in the southern hemisphere and the uncertainty ellipse overlaps with that of the mean pole from this study (Fig. \ref{fig:EI_results}. However, their mean pole developed from red beds near Sault Ste Marie is in the northern hemisphere with a latitude of 12\textdegree N (Fig. \ref{fig:EI_results}). Despite the large uncertainty ellipse associated with this mean pole position, it is distinct from the mean pole position from Jacobsville rocks of the Keweenaw Peninsula and Marquette. Instead, it overlaps with the mean pole position of the ca. 1070 Ma Upper Freda Formation (Fig. \ref{fig:EI_results}; \citeA{Henry1977a}). We hypothesize that the red beds collected by \citeA{Roy1978a} in Sault Ste Marie area are likely Freda-equivalent sediments, instead of being a part of the Jacobsville Formation.

Deposition of the Bayfield Group sediments west to the Keweenaw Peninsula has been hypothesized to be coeval with the Jacobsville Formation but chronological constraints have been lacking. Our new paleomagnetic data from the Jacobsville Formation reaffirms that the pole positions of the the Chequamegon Sandstone of the Bayfield Group and the Jacobsville Formation are close albeit distinct (Fig. \ref{fig:EI_results}). Given that the paleomagnetic pole from the Chequamegon Sandstone of the Bayfield Group (pole longitude=177.7\textdegree E, pole latitude=12.3\textdegree S, A95=4.6\textdegree; \citeA{McCabe1983a}) was calculated without applying inclination correction, the two formations could have even closer a temporal association than the current paleomagnetic pole configuration appears (Fig. \ref{fig:EI_results}). In addition, the fact that only reversed polarity directions were recovered from the Chequamegon Sandstone could indicate that deposition of the sandstone collected by \citeA{McCabe1983a} could be coeval with one of the reversed polarity period recorded by the Jacobsville Formation (Fig. \ref{fig:strat_column}). To further test the temporal correlation between the sedimentary units, detailed paleomagnetic study and detrital zircon U-Pb geochronology data from the Chequamegon Sandstone are needed.

\subsection*{Paleogeography of Laurentia in the late Proterozoic and the development of supercontinent Rodinia}

Laurentia's apparent polar wander is well constrained in the Mesoproterozoic by the Keweenawan Track thanks to the large number of paleomagnetic poles paired with high-precision geochronology from rocks of the ca. 1.1 Ga Midcontinent Rift \cite{Swanson-Hysell2019a}. In addition, the Midcontinent Rift have had simple a thermal history for their old ages, as rifting ceased prior to lithospheric separation, preserving rocks of the rift within the continental interior far from continental margin and subsequent orogenesis. Since the end of the Midcontinent Rift mamatism, igneous paleomagnetic records from the the interior of Laurentia is lacking, and data from rocks associated with the ca. 1080 to 980 Ma Grenville Orogeny have been used to infer for Laurentia's plate configuration in the late Mesoproterozoic to early Neoproterozoic \cite<e.g.>{Weil1998a, Halls2015b}. However, in contrast to rocks of the Midcontinent Rift which usually have straightforward paleomagnetic records \cite<e.g.>{Swanson-Hysell2019a}, rocks of the Grenville Province which experienced up to granulite facies metamorphism have much more complicated thermal histories and the timing of their magnetic remanence acquisition and alteration is more difficult to constrain. As is shown in Fig. \ref{fig:pole_plot}, paleomagnetic poles developed from rocks of the Grenvillian Orogeny plots near Australia with respect to today's continental configuration, forming arc distances ranging from $\sim$35\textdegree to more than 50\textdegree away from the pole of the ca. 1070 Ma Freda Formation \cite{Henry1977a}. Such large discrepancies between pole positions associated with uncertainties in the age of the magnetization of the Grenville rocks led to the controversial ``Grenville Problem"---did the Grenville rocks acquire magnetic remanence prior to the continental collision and thus the pole discrepancies are a result of significant crustal shortening (i.e. the ``two plate model"; \citeA<e.g.>{Irving1972a, Palmer1973a, Buchan1973a, Halls2015b}), or did they acquire remanence after the continental accretion and thus the southerly poles from the Grenville rocks represent Laurentia's continued motion through more recent times (i.e. the ``one plate model"; \citeA<e.g.>{Roy1978a, McWilliams1975a, Weil1998a})?

The two plate hypothesis is based on the interpretation that timing of magnetic remanence acquisition of the Grenville rocks was close to that of the Midcontinent Rift rocks \cite<e.g.>{Warnock2000a, Halls2015b}, but the lack of southerly paleomagnetic poles from the interior of Laurentia is interpreted to indicate that the Grenville rocks acquired remanence on allochthonous terranes before the final accretion to Laurentia craton. Thus, up to 4000 $\pm$ 1000 km of crustal shortening was invoked by \citeA{Halls2015b} to explain the large arc distance between the poles.
